---
title: "Design Notes: sql.template"
author: "Christopher Brown"
date: "December 23, 2015"
output: html_document
---

SQLTemplate is an application of 
[Mustache](https://mustache.github.io/)"logic-less templates" to Structured 
Query Language (SQL) with a few added added enhancments. In 
additions to the benefits of Mustache, SQL templates provides a features to 
(de)activate parts of a SQL Template. This features promotes the query reuse and 
sharing by different users, programs, languages and applications.  

## Features

SQLXXX has the following FEATURES.

- Files are:
  - plain text
  - valid SQL -- run as-is, as a default query
- Multi-purpose; simple logic allow for use from multiple languages/applications
- Portable to different application and languages that use the SQLXXX syntax
  - Reduce inconsistencies across applications
- Patterned after existing standards
  - SQL 
  - [Mustache](https://mustache.github.io/)
- STRICT or PERMISSIVE: can warn or raise errors if constraints are not supplied

  
## Key Functions 

 - sql_read : Read SQL statement from a connection or string
 - sql_render : like whisker render, parse the template and substitute values 
   from key-value pairs 
 - sql_activate : Activate (uncomment) part(s) of a SQL template
 - sql_strip_comments : Strip comments from a SQL STATEMENT


## Background

Strictly speaking any SQL template solution needs to provide several 
capabilities:

- parameter substitution 
- variable checking 
- conditional logic / query activation ...
  - e.g. only print if 
  - ? Only print line if variable substitutes
  - Default is strict: check vars / prior to substitutubs 


### Parameter Subsitution

Mustache (via the whisker/whisker.tools packages) is the easiest and probably 
most ubiquitous way to do parameter substitution.

    SQL Template:
        SELECT * 
        FROM table 
        WHERE name = {{name}}
     
    Hash:
        {"name": "foo"}

Parameter substitution will not work if: 
- foo valid: performs substition
- foo missing: raises error
- foo FALSE: 

 

```{r}
# These are really tests 
tmpl <- "SELECT * FROM TABLE WHERE name = '{{name}}'"
data <- fromJSON('{"name": "foo"}')

data1 = list( name = "foo" )
data2 = list( name = FALSE )
data3 = list() 
data4 = list( other_name = "roger") 

whisker.render(tmpl,data1)  # Shown  # Works 
whisker.render(tmpl,data2)  # FALSE not-shown
whisker.render(tmpl,data3)  # DOES not-sexist
whisker.render(tmpl,data4)  # No substitution

```


## (Un)Masking, unmasking, activating 

The second part of a sql template needs to support conditional logic, namely 
turning on or off logic.  This can surprisingly be done with 

To get whisker to work as a template engine, we need to have a convention for using 
comments to indicate 

    Template:
      SELECT
        A  
        --#foo:, B  -- RENDERS IF DEFINED OTHERWISE NOT
      FROM table 1
     
    Hash:
       { "foo" : true }   

  

Here, two things special are done.
```{r}
tmpl <- "
      SELECT
        A,  
        --foo: B,  -- RENDERS IF DEFINED OTHERWISE NOT
      FROM table 1
"

sql.template::sql_render( tmpl, list(foo=TRUE) )

```

## Poor interface

*sql.template* is poorly designed. It requires two arguments to *sql_render* to render text:

* `tags = ...` determine what is uncommented 
* `data` determines that is rendered

It may be possible to perform both in one step by using only the `data` argument. If the value is exists/than the line should be uncommented.


# Order of operations

Current: Uncomment, substitute, strip ...

Proposed: Preform all substitutions first! 
* substitute 
* uncomment 
* strip remaining comments (optional)

# Functionality

sql.template provides for the following functionality:

* Uncommenting, especially 
** SELECT clause to select more columns 
** WHERE clause to filter rows

* Parameter substitute using mustache

* BOTH 


## Uncommenting 

Uncommenting can be activated if/when a value exists in data. For example:

  - Evaluate a condition in line to true in 
    /*{{TRUE}}: some sql */   --> some sql  
    --{{TRUE}}: some sql      --> some sql 
    
    If whisker.render if the variable does not exist, nothing is inserted 
    --r: --> --: (which is a nice way to indicate that nothing was substituted.)
    
Parameterized comments are uncommented if the value is evaluated.      

### Uses separet parameters:    

    /*{{filter_a}}: AND a = {{a}} */
    --{{filter_b}}: AND b = {{b}}
    
    
    data=list( filter_a=TRUE, filter_b=FALSE )
    

# Using the parameter alone 

This is the perfered method

    --{{r}}: AND r = {{r}}
    /*r: ... */ 
    
    data = list( r = 1 )
    
NOTE: list( r=FALSE ) will also get rendered.


## Required Changes

### Order of operations

Currently, sql.template uncommented then substitutes. This will have to change. 

    
   
    
## Alternatives

* No parameters, parameter sets/unsets the SQL statement

    SELECT 
        name
      /**{{TRUE}}, date**/
    FROM TABLE 
    
    WHERE 
      1 = 1 
      AND NAME = {{NAME}}

* Parameter can be filled -> gets filled 
* Multiple parameters defined.


### Most desirable

The most desirable situation is to have a SQL parser that can turn a basic SQL
statement into a code structure that can subsequently get modified. 

   "Select * from table" %>% where( name = "Fred" )
   
I don't believe that `dplyr` provides this interface. *sql.template* uncommenting is a work-around this feature.    


# DPLY vs SQLTemplate

One comment that is often made is why not just use dplyr. To understand the 
rational, it is important to make some distinction between dplyr and 
SQLTemplate.

|    dplyr                         |    SQLTemple                 |
|----------------------------------|------------------------------|
| Build query using logic pipeline | Build Query in text template |
|-----------------------------------|------------------------------|
| Non-transportable code (as of now)| Language/Application agnostic 
|                                   |  just a standard             |

## DPLYR

- Build your query through a logic pipeline.
- Non-transportable code (as of now)

## SQLMustache

- Language Independent
- logic-late
- Runs and parses as-is. 

    SELECT 
      name 
      --addr: addr 
    FROM 
      person
      --addr: INNER JOIN address on personId
    WHERE 
      1=1 
      --addr: AND addr like '%MOCKING BIRD%' 
  

# TRUTHY VALUES exists, ! NULL, is not empty or NA
